---
title: Peer Assessment 2 - Population Health and Economic Consequences of Severe Weather
  Events in the US
output: pdf_document
---

## synopsis 
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities.
Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent 
possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This 
database tracks characteristics of major storms and weather events in the United States, including when and where they 
occur, as well as estimates of any fatalities, injuries, and property damage.

The objectives of this analysis is to answer the question of 1. Across the United States, which types of events (as 
indicated in the EVTYPE variable) are most harmful with respect to population health? 2. Across the United States, which 
types of events have the greatest economic consequences?

```{r setup, include = FALSE}
require(plyr)
require(ggplot2)
```

## Data Processing
### Loading data
The data got downloaded from [source](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and stored as `stormData.csv` file at the disk in the current working directory.

```{r cache=TRUE}
if (!file.exists("repdata_data_StormData.csv")) {
    # Download the file if it is not represented in the working directory
    if (!file.exits("repdata_data_StormData.csv.bz2")) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
            "repdata_data_StormData.csv.bz2", mode = "wb")
    }

    # unzip
    bunzip2("repdata_data_StormData.csv.bz2", exdir = ".")
}

storm <- read.csv("repdata_data_StormData.csv")
```
####The dataframe has 902,297 observation and 37 variables.
```{r}
dim(storm)
names(storm)
```
### Preprocessing

To find the event types that are most harmful to population health and have the most economic consequences, the number
of casualties and property damages are aggregated by the event type, separately.
```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
fatalities = sum(FATALITIES))
# Find events that caused most death and injury
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
fatal_events[, c("EVTYPE", "fatalities")]
```
Top 10 events that are most harmful to population health (I use deaths as the indicator)
```{r}
fatal_events[, c("EVTYPE", "fatalities")]
```

```{r}
propdmg <- ddply(storm, .(EVTYPE), summarize,
propdmg = sum(PROPDMG))
# Find events that caused most death and injury
propdmg_events <- head(propdmg[order(propdmg$propdmg, decreasing = T), ], 10)
```
Top 10 events that have the most Economic Consequences (I use property damages as the indicator)
```{r}
propdmg_events[, c("EVTYPE", "propdmg")]
```


### Results
```{r}

names <-fatal_events$EVTYPE
barplot(fatal_events$fatalities,main="Fatality of Different Severe Weather Events", horiz=FALSE,names.arg=names, las=2)
```



```{r}

names <-propdmg_events$EVTYPE
barplot(propdmg_events$propdmg,main="Property Damages of Different Severe Weather Events", horiz=FALSE,names.arg=names, las=2)
```

## Conclusions
Based on the data analysis tornadoes were the most harmful severe weather events, in terms of casualties and economic consequences in the last 60 years. Excessive heat and flash flood are important cusses for casualties. Heat is not an importance factor for property damages.  Overall, tornados are the most harmful event for both human health and economic damages.  Too much water (flash flood, flood) and high wind (thunderstorm wind and wind) have the most economic consequences.  Lightning is also an important severe event for both human health and property damages.
